A sustainable approach to tourist signage on (historical) trails

### 2023-07-20 paper.tex
Prima revisione quasi competa, aggiunte immagini, tabella e mappa tag, eliminato frammento duplicato

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

@@

\subsection{The role of tourist signage}

\section{Tourist signage sustainability}

\subsection{Working hypothesis}

\subsection{Evaluation parameters}

\subsection{Evaluation of sample alternatives}

\section{Application to a concrete use case}

\subsection{Trail selection}

\subsection{Social and political aspects}

\subsection{Signage design}

\subsection{Monitoring results}

\section{Conclusions}

\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}

The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

@@

This paper provides a concept solution of a specific aspect. The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding. Given the tighly interconnected nature of the situation, we observe how a vary specific aspect, the touristic signage, exhibits effects on the society, economicy and environment.

@@

Its functions are twofold
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point to this end, but not all visitors feel comfortable reading a map. The ideal signage should include visible reference points and explicit actions, like {\em turn right after crossing the stream}. A good example of such a signage is the one used by the Italian Alpine Club (CAI): it consist of a colored stripes marked on tree bark. The marks are placed in such a way that, standing nearby one of them, the hiker is able to see the successive on the trail
\item a sign should explain the reasons of interest of a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit. In this sense, the CAI signal above is not of use. A simple board with a site name may be sufficient only when the site is sufficiently reknown. Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
\end{itemize}

@@

The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.

A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:

@@

\subsection{Technology to minimize intrusion}

We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the habitat, but the impact of such intervention must be better than that related with traditional signs. To this end we include in our solution tools that are not part of the site but remain with the user.

During the last decades, we witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a trend. Smartphones empower individual communication capabilities, allowing to receive sound, visual and tactile interactions. The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}. to investigate ways to exploit such tools in our context. Also in this case we start from the limits of such technology, especially in the case of outdoor activities.

@@

@@

Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.

@@

Circular economy

@@

An exhaustive solution is described by P. Liu in \cite{liu16a}, which provide details for an infrastructure that guides the visitors inside an urban milieu. In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on communication.

Wan \cite{wan22a} evaluates the quality signage, without making reference to a specific technology, but with many example showing phisical boards, using a formal reference, the Universal Design Principles \cite{udi97a}.

The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism. The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to a vast literature.

A relevant research direction covers the social aspects of related with the use of the smartphone. W. Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which again is considered in the Internet despite the title indicates a nature-based destination, where notably the Internet is not always reachable.

@@

In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences. The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the imact on the social fabric.

The present paper wants to fill this gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support taking into account sustainability by design. Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.

We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would represent the success of a strategy deserve further investigation. We aim at isolating an issue, propose a strategy for its solution and implement a proof of concept for it.

@@

A straightforward approach concentrates on the wireless capabilities of such a device. Given the premise that the Internet is not reachable, the host might provide a local network of low power radios covering the region of interest. Small servers connected to the network might provide specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage fo tens of meters, and a volume in the order of the $cm^3$. They are able to provide WiFi Access Points as well as Web contents. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

The dependency from power supply is a serious limitation of such kind of solution. A radio transmitter is a rather power consuming device, in the range of Watts. Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organization should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device.

@@

An alternative consists in the utilization of passive devices, like Near Field trasmitters. The device is flat with a surface ranging from a credit card to a coin. The receiver must be very close to enable the device to transmits its content. The power needed to operate the radio is drained from the smartphone, so batteries are not needed. The capacity is in the range of the KB, one page on this journal.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the smartphone as a chink of text, and can be treated as such. The device can read it aloud to meet the user's inabilities or a translator. For both functions off-line apps exist.

Another solution in the family of passive devices is the QR-code. Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text. The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.

@@

\begin{itemize}
	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single touch
	\item{\em Perceptible information} is a critical requirement, which contrasts with the aim of keeping low the environmental intrusion. This point will be further discussed in the section devoted to the implementation
	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution
	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
	\item{Size and Space for approach and use} have to be carefully considered. In the case of the NFC the smartphone needs to be nearly in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
\end{itemize}

Such a minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares well compared with other solutions that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit. If capacity is not an issue a solution based on NFC or QR-codes exhibits a number of advantages:
\begin{itemize} 
\item does not require any power supply
\item has a limited impact on the landscape
\item is not exposed to theft
\item has a negligible cost
\item is durable
\item produces a limited quantity of waste when disposed
\item the content can be stored by the user's device for successive utilization
\end{itemize}

@@

A solution based on the user smartphone associated with passive devices compares favorably with traditional signage boards. Considering the limits of the latter approach listed in table \ref{tab:board}:

@@

There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among the

@@

In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit. To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.

The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.

A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:

\begin{itemize}
\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
\item update limits: to update the content, the plaque must be replaced
\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
\end{itemize}

During the last decades we have witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a thread. Smartphones empower the communication capabilities of an individual, allowing to receive sound, visual and tatcile interactions. It is therefore straightforward to consider the way to exploit such capabilities in our context. Also in this case we start from the limits of such technology.

Concentrate on outdoor activities

One is related with the diffusion of such a technology, which, although very diffused, is not equally familiar to everybody. The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure.

Starting from the two aspects above we envision the directions towards a successful strategy, keeping in mind the overall sustainability of the approach.

Regarding usability, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.

The second issue is more complex, especially when considering outdoor activities. In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact. Consider the installation of antennas to cover a wide area and power supply for the radios.

Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.

One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region. We consider the implementation of several small servers operating in isolation that provide a few HTML pages. The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$. They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.

The dependency from power supply is a serious limitation of such kind of solution. A radio transmitter is a rather power consuming device, in the range of Watts. Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organizaton should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device,

An alternative consists in the utilization of passive devices, like Near Field trasmitters. The device is flat with a surface ranging from a credit card to a coin. The receiver must be very close to enable the device to transmits its content. The power needed to operate the radio is drained from the smartphone, so batteries are not needed. The capacity is in the range of the KB, one page on this journal.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is trasferred to the smartphone as a chink of text, and can be treated as such. The device can read it aloud to meet the user's inabilities or a translator. For both functions off-line apps exist to compensate the lack of network connection.

Another solution in the family of passive devices is the QR-code. Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that can be translated into a text. The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera. We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.

Such a minimalistic solution compares surprisingly well when compared with other solutions that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words are sufficient to convey the description of a site, or provide directions for the the visit. In exchange, the QR-code solution does not require any power supply, is extremely easy to use, has a limited impact on the laandscape, is not exposed to theft, has a negligible cost, is durable and produces a limited quantity of garbage when disposed, is readable at a moderate distance and can be read aloud, translated or stored by the user's device. It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.

The adoption of QR-codes compares favorably with traditional boards, considering the strengths and limits as listed above.

\begin{itemize}
\item dimensions: a 2-300 characters QR-code has the dimensione of a poker card
\item logistics: QR-code board can be installed on any sort of preexisting support
\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
\item accessibility for the visually impaired: the text can be read aloud
\item international: the text can be translated
\item update: the board can be easely replaced when the content turns obsolete
\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
\end{itemize}

One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha

may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.

One option discriminates the use of an electronic device on the side of the user.

In case we assume the user does not have a electronic device with her, the 
One option is to introduce active electronics: the touristic site administration provides a

>>>>

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}
\usepackage{float}
\usepackage[caption = false]{subfig}

@@

%\setcounter{section}{-1} %% Remove this when starting to work on the template.
%\section{How to Use this Template}

%The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

@@

This paper provides a concept solution of a specific aspect. The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding. Given the tighly interconnected nature of the situation, we observe how a very specific aspect, the touristic signage, exhibits effects on the society, economy and environment.

@@

The functions of signage are twofold
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point to this end, but not all visitors feel comfortable reading a map. The ideal signage should include visible reference points and explicit actions, like "turn right after crossing the stream". A good example of such a signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark. The marks are placed in such a way that, standing nearby one of them, the hiker is able to spot the successive on the trail
\item a sign should explain the reasons of interest of a site. This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit. In this sense, the CAI signal above is not of use. A simple board with a site name may be sufficient only when the site is sufficiently reknown. Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
\end{itemize}

@@

%In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit. To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.

The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible. A traditional approach makes use of physical information boards with graphical or textual contents.

%The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.

Our study starts from pointing out the issues related with such a solution:

@@

%A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:

%\begin{itemize}
%	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
%	\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
%	\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
%	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
%	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
%	\item update limits: to update the content, the plaque must be replaced
%	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
%\end{itemize}

%Such considerations motivate an interest for an alternative way of communication.

\subsection{Technology to minimize intrusion \label{sec:minimize}}

We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the environment, but the impact of such intervention must be better than that related with traditional signs. To this end we include in our solution tools that are not part of the site but remain with the user.

During the last decades, we witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a trend. Smartphones empower individual communication capabilities, allowing to receive sound, visual and tactile interactions. The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}. to find ways to exploit such tools in our context. Also in this case we start from the limits of such technology, especially in the case of outdoor activities.

@@

% The second issue is more complex, especially when considering outdoor activities. In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact. Consider the installation of antennas to cover a wide area and power supply for the radios.

@@

Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of familiarizing with new applications.

@@

% Circular economy

@@

An exhaustive solution is described by P. Liu in \cite{liu16a}, which details an infrastructure that guides the visitors inside an urban milieu. In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on Internet connectivity.

Wan \cite{wan22a} evaluates the quality of signage, without making reference to a specific technology, but with many examples showing physical boards, using as a formal reference the Universal Design Principles \cite{udi97a}.

The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism. The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to the vast literature.

A popular research direction covers the social aspects related with the use of the smartphone. W. Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered in the Internet, despite the title indicates a nature-based destination, where notably the Internet is not always reachable.

@@

In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences. The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the impact on the social fabric.

The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined tourism support aspect, taking into account by design its sustainability. Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.

We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve further investigation. We aim at isolating an issue, proposing a strategy for its solution and implementing a proof of concept for it.

@@

A straightforward approach concentrates on the wireless capabilities of such a device. Given the premise that the Internet is not reachable, the {\em host} stakeholder may provide a local network of low power radios covering the region of interest. Small servers connected to the network might provide specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$. They are able to provide WiFi Access Points as well as Web contents. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

% One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region. We consider the implementation of several small servers operating in isolation that provide a few HTML pages. The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$. They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.

Power supply dependency is a serious limitation for such kind of solution. A radio transmitter is a rather power consuming device, in the range of Watts. Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organization should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device.

@@

An alternative consists in the utilization of passive devices, like Near Field trasmitters. The transmitting device is flat with the size of a coin and a cost of less than one dollar per piece. The receiving smartphone must be very close to receive the content. The power needed to operate the radio is drained from the smartphone, so that batteries are not needed. The NFC device capacity is in the range of the KBytes, nearly a page on this journal. Its manufacturing requires a smartphone with a specific app.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text which can be treated as such. The smartphone can read it aloud to compensate user's inabilities or a translator to cope with linguistic issues without need of an Internet connection.

Another solution in the family of passive devices is the QR-tag. Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text. The capacity of a QR-tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR-code roughly equals that of an NFC tag. A QR code is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.

@@

\begin{itemize}
	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows listening to the info instead or reading, translating the information in a different language, or storage for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single touch
	\item{\em Perceptible information} is a critical requirement, which contrasts with the aim of keeping low the environmental intrusion. This point will be further discussed in the section devoted to the implementation
	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution
	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
	\item{Size and Space for approach and use} have to be carefully considered. In the case of the NFC the smartphone needs to be nearly in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
\end{itemize}

Such minimalistic solutions (see figure \ref{fig:qrnfccoin}) compares well with other solutions that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit. If capacity limits are not an issue, a solution based on NFC or QR-codes exhibits a number of advantages:
\begin{itemize} 
\item does not require any power supply
\item has a limited impact on the landscape
\item does not entice theft
\item has a negligible cost
\item is durable
\item produces a limited quantity of waste when disposed
\item content can be stored for usage when the user reaches a zone covered by the Internet
\end{itemize}

The two passive technologies of choice exhibit the following features, that may make one of them preferable for a specific application:

\begin{itemize}
	\item an NFC-tag is smaller than a QR-tag
	\item writing an NFC-tag requires a smartphone, while the QR-tag needs a printer
	\item reading an NFC-tag operates at near-to-contact distances, while QR-codes can be read from meters away
\end{itemize}

@@

For a signage application the QR-tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalisms and misuse.

In conclusion, we have reasons to select a QR-code based solution as a good candidate for a smartphone-based signage. We now consider how it it copes with the limits of a traditiona board-based approach (as listed in table \ref{tab:board}):

@@

There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among visuali impact, and visibility.

Il secondo?

@@

As anticipated in the introduction, our study wants to provide a proof of concept for a sustainable solution in a concrete setup. To comply with such an holistic approach, we need to start from the definition of the operation context.

This section is devoted to the description of the economical and social context where we are going to operate, as well as the historical backgound representing the heritage resource we want to promote. Next we will discuss the technical detail of the solution.

The geographical area of interest are the surroundings of Casoli, a small village in a mountaineous region in the north of Tuscany, in Italy. The area fall within the municipailty of Bagni di Lucca, in the province of Lucca. The archaeological team of our project formerly realized the municipal archaeological map in 2021 \cite{letizia2021}, thus assessing the reasons of interest for the heritage and the geomorphological features of the area.

\subsection{The natural and cultural resources of Casoli and its vicinity}

Bagni di Lucca is located on the north-eastern boundary of the Province of Lucca, in Val di Lima, and is part of the Media Valle del Serchio district. With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological and historical heritage, both expressed and as yet unexpressed. Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are catalogued, photographed and georeferenced.

From a historical-geographical point of view, this area is identified with the Lima stream and its tributaries, that impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.

The Lima stream has characterised the history of Bagni di Lucca since ever: some of the oldest evidence of human presence in the valley has
been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the
manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited the since the Middle Age until the 1980s \cite{men76, bed05, ser21}.

On the naturalistic side, the region of Bagni di Lucca encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{natura2000} European Economic Community (EEC) initiative.

There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the Lima stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
\begin {itemize}
\item the limestone areas of Val di Lima and Balzo Nero; 
\item Monte Prato Fiorito-Monte Coronato-Valle dello
Scesta; 
\item Orrido di Botri.
\end{itemize}

\begin{figure}
	\centering
	\includegraphics[width=0.3\linewidth]{figure/toscana-casoli}
	\caption[The region of Casoli]{}
	\label{fig:toscana-casoli}
\end{figure}

The latter is also a SPA (Special Protection Area) and the Orrido di Botri State Reserve is located within it \cite{nat00}. It is therefore a natural heritage with a fragile balance that needs to be preserved.

In recent years, the main tourist attractions have concentrated on the Lima stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting and sup. Others important tourist attraction are trekking and hiking, supported by a network of path tracked by local CAI (Club Alpino Italiano).

In recent years the community opened two entirely new trails:
\begin{itemize}
	\item the Alta Via dei Pastori (2019), a ring around Monte Prato Fiorito that takes up the ancient grazing route, and
	\item t the Sentiero degli Avi (2020), a ring that from Montefegatesi
	reaches Monte Coronato \cite{pin21}
\end{itemize}

Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of Pistoia, with a
variant that from Popiglio continues in five stages in the municipality of Bagni di Lucca to Pieve di Controne, acting as the 'Lucca gateway' to the Path \cite{camsb}.

Such initiatives led to a considerable revival of interest among Italian and international hikers for the area, especially for the Apennine side of the valley.

We see the opportunity to improve the valorisation and enjoyment of this area by applying a slow tourism approach and involving the community, especially in the southern part of the municipality, which is still little frequented, more hilly and therefore less travelled by the trail network. We propose the creation of geo-itineraries characterised by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.

\subsubsection{An historical perspective of an italian mountain site}

% The village of Casoli is located on the Lima stream. To the south-east of the settlement lies a modest relief with a small glacial lake. The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age . These frequentations can be attributed to a seasonal exploitation of resources and, later, with the climatic improvement, as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

South-east of the today location of Casoli lies a modest relief and a small glacial lake. The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

Ceramic fragments dated between the 3rd century B.C. and the 1st-2nd century A.D. prove that Ligurian populations occupied the region scattered and in small nuclei. The dedication of the Latin colony of Lucca in 180 B.C. marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

We have little evidence of settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.

% Just outside the Casoli settlement, ceramic fragments datable between the 3rd century B.C. and the 1st-2nd century A.D. emerged from an excavation for the construction of a bowling green, testifying, together with other documented evidence in the area of Bagni di Lucca, how Ligurian populations settled in Val di Lima at least from the 3rd century B.C. onwards.

% In this period, the settlement must have been rather scattered and in small nuclei. The dedication of the Latin colony of Lucca in 180 B.C. marked a decisive turning point in the Romanisation of the area and within a few years there was a definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

% From the data currently available, it can be stated that the mountainous hinterland retained a relatively marginal role. Archaeological traces for the Roman era are scarce, apart from the material that emerged from Buca La Piella, located east of Casoli along the Solco del Monte.

%The cave was discovered and investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.

% This confirms a rarefied or seasonal frequentation with a possible division in “fundi”, probably linked to the exploitation of the mountain's natural resources, the products of the forest and pasture, as well as having represented a connection route with the Po Valley area, through the Apennine passes \cite{gia96, men81, cia03}.

In addition to the scarcity of archaeological remains, there is a profound silence in the historical and literary sources for the Val di Lima from Roman and Late Antique times, although it can be assumed that the Val di Lima followed the fate of the city of Lucca, first with the occupation of the Goths and the brief Byzantine rule and then with the Longobard invasion and the formation of the Duchy of Lucca. In the hinterland, the Lombard settlement was concentrated according to a system consisting of castles and villages, mostly evidenced by documents from the 8th and 9th centuries and the remains of burial areas. In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.

The only find from this period is the Grotta di Arzale, also known as the 'Grotta Murata', a rock shelter that opens up in the locality of Sevilucchio, north-west of Casoli along a historic road connecting Cocciglia to Casoli. Fictile artefacts, defined as being from the 'barbarian' era, have been found here; in reality, the bibliographic and oral sources collected have indicated fragments of an olla with threading, the attestation of which in the archaeological contexts of Lucca and northern Tuscany starts from the 6th-7th century and may reach as far as the end of the 11th century \cite{gia96}.
